Normalization by Evaluation for Sized Dependent Types

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Moved equality rules to own file.
[ ICFP 17 ] Moved typing rules etc. to own file.
\tau instead of tau
[ ICFP 17 ] Space correction after if.

For concrete function types and terms, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
$\forall i.\ (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
% $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstraction takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
We may also use names when we construct concrete contexts, for instance, if $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
$\Gamma.\Size.\Nat \ind 0$.

@@

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory.

@@

Inference rules.

\fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
\[
  \ru{}{\der \cempty}
\qquad
  \ru{\der \Gamma \qquad
      \resurrect \Gamma \der T
    }{\der \cext \Gamma T}
\qquad
  \ru{\der \Gamma
    }{\der \cext \Gamma \Size}
\qquad
  \ru{\der \Gamma
    }{\der \erext \Gamma \Size}
\]

\fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$).
\[
  \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
\qquad
  \ru{\Gamma(i) = \eann T
    }{(\cext \Gamma \_)(i+1) = \eann T \slift}
\]

\fbox{$\Gamma \der a : \Size$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der a : \Size$).
\[
  \ru{\der \Gamma}{\Gamma \der \infty : \Size}
\qquad
  \rux{\der \Gamma}{\Gamma \der o : \Size}{o \in \NN}
\qquad
  \ru{\der \Gamma \qquad \Gamma(i) = \erann[\noterased]\Size
    }{\Gamma \der \ind i + o : \Size
    }
\]

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : \Set[0]}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{%\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der a, b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\resurrect\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      %\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der a,b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (a + 1) \qquad
      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, (\zero{a}) \qquad
      \Gamma \der t_s : (x : \Nat a) \to T \, (\suc{a} x)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat a \qquad
      \Gamma \der T : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
\qquad
  \ru{\Gamma \der t : T \qquad
      \resurrect\Gamma \der T \leq T'
     }{\Gamma \der t : T'}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

\fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)

Computation rules.
\begin{gather*}
  \ru{\cext \Gamma U \der t : T \qquad
      \Gamma \der u : U
    }{\Gamma \der (\lambda t)\,u = t[u] : T[u]}
\qquad
  \ru{\cext \Gamma \Size \der t : T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der (\lambda t)\,a = t[a] : T[a]}
\qquad
  \ru{\erext \Gamma \Size \der t : T \qquad
      % \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der a, b : \Size
    }{\Gamma \der (\lambda t)\,\ann a = t[a] : T[b]}
\\[2ex]
  \ru{\resurrect\Gamma \der a, b : \Size \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      % \Gamma \der a : \Size \qquad
      \Gamma \der t_z : T\,\zero{b} \qquad
      \Gamma \der t_s : (x : \Nat b) \to T\,(\suc{b}x)
    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{b}}
\\[2ex]
  \ru{\resurrect\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T\,\zero{b} \qquad
      \Gamma \der t_s : (x : \Nat b) \to T\,(\suc{b}x)
    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{b} t)}
\\[2ex]
  \ru{\Gamma \der c : \Nat b \qquad
      \resurrect\Gamma \der a : \Size \qquad
      \Gamma \der T : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\,T %% \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
     }{\Gamma \der c\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, c : T \,b \, c
     }%{c \in \{\zero a, \suc a t'\}}
\end{gather*}
Extensionality rules.
\begin{gather*}
  \ru{\Gamma \der t : \piT U T
    }{\Gamma \der t = \lambda x. t\,x : \piT U T}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \cext {\resurrect\Gamma} \Size \der a : \Size
      %\erext \Gamma \Size \der a : \Size
    }{\Gamma \der t = \lambda x. t \ann a : \forallT T}
\end{gather*}
Congruence rules.
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] = \Set[\ell]: \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U = U' : s \qquad
      \cext \Gamma U \der T = T' : s
    }{\Gamma \der \piT U T = \piT{U'}{T'}: s}
\qquad
  \ru{\cext \Gamma \Size \der T = T' : s
    }{\Gamma \der \epiT \Size T = \epiT \Size T' : s}
\\[2ex]
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a = \Nat a : \Set[0]}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i = \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t = t' : T
    }{\Gamma \der \lambda t = \lambda t' : \epiT U T}
\qquad
  \ru{\Gamma \der t = t' : \piT U T \qquad
      \Gamma \der u = u' : U
    }{\Gamma \der t\,u = t'\,u' : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\\[2ex]
  \ru{\Gamma \der t = t' : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a = t'\,a : T[a]}
\qquad
  \ru{\Gamma \der t = t' : \forallT T \qquad
      % \Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der a,a',b : \Size
    }{\Gamma \der t \ann a = t' \ann{a'} : T[b]}
\\[2ex] % \qquad
  \ru{%\Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der a, a', b : \Size
    }{\Gamma \der \zero a = \zero {a'}: \Nat (b + 1)}
\qquad
  \ru{\resurrect \Gamma \der a, a' : \Size \qquad
      \Gamma \der t = t' : \Nat b
    }{\Gamma \der \suc a t = \suc {a'} t' : \Nat (b + 1)}
\\[2ex]
  \ru{\Gamma \der u = u' : \Nat (a + 1) \qquad
      \Gamma \der T = T' : \Nat (a + 1) \to \Set[\ell] \\
      \Gamma \der t_z = t_z' : T \, (\zero a) \qquad
      \Gamma \der t_s = t_s' : (x : \Nat a) \to T \, (\suc a x)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s = u'\,\case[\ell] {T'} {t_z'} t_s' : T \, u}
\\[2ex]
  \ru{\Gamma \der u = u' : \Nat a \qquad
      \Gamma \der T = T' : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t = t' : \FixT\,T \qquad %% \allT i \Size (\funT x {\Nat i} T \,i\, x) \to \funT x {\Nat (i + 1)} T \,{i+1}\, x
     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \, a \, u}
\qquad
  \ru{\Gamma \der t = t' : T \qquad
      \resurrect\Gamma \der T \leq T'
     }{\Gamma \der t = t' : T'}
\end{gather*}
Equivalence rules.  (Reflexivity is admissible.)
\begin{gather*}
  \ru{\Gamma \der t = t' : T
    }{\Gamma \der t' = t : T}
\qquad
  \ru{\Gamma \der t_1 = t_2 : T \qquad
      \Gamma \der t_2 = t_3 : T
    }{\Gamma \der t_1 = t_3 : T}
\end{gather*}

\fbox{$\Gamma \der T \leq T'$} (implies $\Gamma \der T,T'$)
% (implies $\resurrect \Gamma \der T,T'$)
\begin{gather*}
  \ru{\der \Gamma \qquad \ell {\leq} \ell'
    }{\Gamma \der \Set[\ell] \leq \Set[\ell']}
\qquad
  \ru{\Gamma \der a \leq b : \Size
    }{\Gamma \der \Nat a \leq \Nat b}
\qquad
  \ru{\Gamma \der T = T'
    }{\Gamma \der T \leq T'}
\\[2ex]
  \ru{\Gamma \der U' \leq U \qquad
      \cext \Gamma {U'} \der T \leq T'
    }{\Gamma \der \piT U T \leq \piT {U'} {T'}}
\qquad
  \ru{\cext \Gamma {\Size} \der T \leq T'
    }{\Gamma \der \epiT \Size T \leq \epiT {\Size} {T'}}
\qquad
  \ru{\Gamma \der T_1 \leq T_2 \qquad
      \Gamma \der T_2 \leq T_3
    }{\Gamma \der T_1 \leq T_3}
\qquad
\end{gather*}

Substitution typing. % and equality.
\fbox{$\Gamma \der \tau : \Delta$}
(implies $\der \Gamma$ and $\der \Delta$
 %% and $\Gamma \der \tau : \resurrect \Delta$  %NO
 [and $\resurrect\Gamma \der \tau : \Delta$]
 and $\resurrect\Gamma \der \tau : \resurrect \Delta$).

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\Gamma \der \sempty : \cempty}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der t : T \tau
    }{\Gamma \der \sext \tau t : \cext \Delta T}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \sext \tau a : \cext \Delta \Size}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \resurrect\Gamma \der a : \Size
    }{\Gamma \der \sext \tau a : \erext \Delta \Size}
\end{gather*}

@@

% Substitution typing and equality.
% \fbox{$\suty \Gamma \sigma \tau \Delta$} and

Substitution equality
\fbox{$\sueq \Gamma \sigma {\sigma'} \tau \Delta$}
(implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \tau : \Delta$ and $\sueq \Gamma {\sigma'} {\sigma} \tau \Delta$).

% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\suty \Gamma \sempty \sempty \cempty}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der u = t : T \tau
%     }{\suty \Gamma {\sext \sigma u} {\sext \tau t} \cext \Delta T}
% \\[2ex]
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a = b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \cext \Delta \Size}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a, b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \erext \Delta \Size}
% \end{gather*}

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\sueq \Gamma \sempty \sempty \sempty \cempty}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der u = u' = t : T \tau
    }{\sueq \Gamma {\sext \sigma u } {\sext {\sigma'} {u'}} {\sext \tau t} {\cext \Delta T}}
\\[2ex]
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \Gamma \der a = a' = b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\cext \Delta \Size}}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \resurrect\Gamma \der a, a', b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\erext \Delta \Size}}
\end{gather*}

\begin{lemma}[Reflexivity]
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = t : T$.
\end{lemma}

@@

\begin{lemma}[Context well-formedness]
\label{lem:cxtwf}\bla
  \begin{enumerate}
  \item If $\der \cext \Gamma \Delta$ then $\der \Gamma$
  \item If\/ $\Gamma \der J$ then $\der \Gamma$.
  \end{enumerate}
\end{lemma}

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

\begin{lemma}[Specific substitutions]
\label{lem:specsub}\bla
\begin{enumerate}
\item If $\der \cext \Gamma \Delta$ then $\cext \Gamma \Delta \der \slift_{|\Gamma|}^{|\Delta|} : \Gamma$.
      If $\der \cext \Gamma T$ then $\cext \Gamma T \der \slift : \Gamma$.
\item If $\der \Gamma$ then $\Gamma \der \sid : \Gamma$.
\item If $\Gamma \der u : U$ then $\Gamma \der \sg u : \cext \Gamma U$.
\end{enumerate}
\end{lemma}

\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}
\begin{corollary}[Partial resurrection for substitution equality]
  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then
  $\sueq {\resurrect \Gamma} \tau \tau \tau {\resurrect \Delta}$.
\end{corollary}

>>>>

For concrete function types and terms, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
$\forall i.\ (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
% $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstraction takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
We may also use names when we construct concrete contexts, for instance, if\/ $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
$\Gamma.\Size.\Nat \ind 0$.

@@

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.

\input{figty}
\input{figeq}

@@

% Inference rules.

% \fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
% \[
%   \ru{}{\der \cempty}
% \qquad
%   \ru{\der \Gamma \qquad
%       \resurrect \Gamma \der T
%     }{\der \cext \Gamma T}
% \qquad
%   \ru{\der \Gamma
%     }{\der \cext \Gamma \Size}
% \qquad
%   \ru{\der \Gamma
%     }{\der \erext \Gamma \Size}
% \]

% \fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$ if\/ $\der \Gamma$).
% \[
%   \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
% \qquad
%   \ru{\Gamma(i) = \eann T
%     }{(\cext \Gamma \_)(i+1) = \eann T \slift}
% \]

%Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

% \fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)

@@

% Substitution typing and equality.
% \fbox{$\suty \Gamma \sigma \tau \Delta$} and
% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\suty \Gamma \sempty \sempty \cempty}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der u = t : T \tau
%     }{\suty \Gamma {\sext \sigma u} {\sext \tau t} \cext \Delta T}
% \\[2ex]
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a = b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \cext \Delta \Size}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a, b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \erext \Delta \Size}
% \end{gather*}

%% COULD BE A LEMMA INSTEAD OF A RULE
% \begin{lemma}[Reflexivity]
%   If\/ $\Gamma \der t : T$ then $\Gamma \der t = t : T$.
% \end{lemma}

@@

\begin{lemma}[Context well-formedness]
\label{lem:cxtwf}\bla
  \begin{enumerate}
  \item If\/ $\der \cext \Gamma \Delta$ then $\der \Gamma$
  \item If\/ $\Gamma \der J$ then $\der \Gamma$.
  \end{enumerate}
\end{lemma}

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if\/ $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

\begin{lemma}[Specific substitutions]
\label{lem:specsub}\bla
\begin{enumerate}
\item If\/ $\der \cext \Gamma \Delta$ then $\cext \Gamma \Delta \der \slift_{|\Gamma|}^{|\Delta|} : \Gamma$.
      If\/ $\der \cext \Gamma T$ then $\cext \Gamma T \der \slift : \Gamma$.
\item If\/ $\der \Gamma$ then $\Gamma \der \sid : \Gamma$.
\item If\/ $\Gamma \der u : U$ then $\Gamma \der \sg u : \cext \Gamma U$.
\end{enumerate}
\end{lemma}

\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If\/ $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If\/ $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If\/ $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}
\begin{corollary}[Partial resurrection for substitution equality]
  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then
  $\sueq {\resurrect \Gamma} \tau \tau \tau {\resurrect \Delta}$.
\end{corollary}

\begin{lemma}[Inversion of typing]
\label{lem:invty} \bla
\begin{enumerate}
\item If\/ $\Gamma \der \Nat a : T'$ then $\Gamma \der a : \Size$ and $\resurrect\Gamma \der \Set[0] \leq T'$.
\item If\/ $\Gamma \der \Set[\ell] : T'$ then $\resurrect\Gamma \der \Set[\ell+1] \leq T'$.
\item If\/ $\Gamma \der \piT U T : T'$ then $\Gamma \der U : s$ and $\cext \Gamma U \der T : s$ and $\resurrect\Gamma \der s \leq T'$ for some $s$.
\item If\/ $\Gamma \der \epiT \Size T : T'$ then $\cext\Gamma \Size \der T : s$ and $\resurrect\Gamma \der s \leq T'$.
\item If\/ $\Gamma \der \ind i : T'$ then $\Gamma(i) = \erann[\noterased] T$ and $\resurrect\Gamma \der T \leq T'$ for some $T$.
\item If\/ $\Gamma \der \lambda t : T'$ then either
  $\cext \Gamma U \der t : T$
  and
  $\resurrect\Gamma \der \piT U T \leq T'$
  for some $U,T$ or
  $\eext \Gamma \Size \der t : T$
  and
  $\resurrect\Gamma \der \epiT\Size T \leq T'$
  for some $T$.
\item If\/ $\Gamma \der t\,u : T'$ then $\Gamma \der t : \piT U T$ and $\Gamma \der u : U$ and $\resurrect\Gamma \der T[u] \leq T'$ for some $U$, $T$.
\item If\/ $\Gamma \der t\,a : T'$ then $\Gamma \der t : \piT \Size T$ and $\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
\item If\/ $\Gamma \der t\,\ann a : T'$ then $\Gamma \der t : \forallT T$ and $\resurrect\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
\item If\/ $\Gamma \der \zero a : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\resurrect\Gamma \der \Nat (b + 1) \leq T'$ for some $b$.
\item If\/ $\Gamma \der \suc a t : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der t : \Nat b$ and $\resurrect\Gamma \der \Nat(b+1) \leq T'$.
\item If\/ $\Gamma \der u \case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
\item If\/ $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
\end{enumerate}
\end{lemma}
\begin{proof}
  Each by induction on the typing derivation, gathering applications of the conversion rule via transitivity of subtyping.
\end{proof}
